IGAS Project

### 2024-04-08 Sections/Appendix/E1_Appendix.qmd
tweaks

```{r}
# print(getwd())
# here::set_here(path='..')
# print(getwd())
source(here::here("Functions", "packages.R"))

>>>>

```{r}
# print(getwd())
# here::set_here(path='..')
# print(getwd())
#source(here::here("Functions", "packages.R"))
pacman::p_load(dplyr,purrr,tidyr,tibble,ggplot2,
  brms,tidybayes, rstanarm,emmeans,broom,bayestestR,
  stringr, here,conflicted, patchwork, knitr, cowplot, RColorBrewer,data.table,gt,
  ggh4x,htmltools,future,furrr,ez)
walk(c("dplyr","here"), conflict_prefer_all, quiet = TRUE)

walk(c("Display_Functions", "org_functions", "Table_Functions"), ~ source(here::here(paste0("Functions/", .x, ".R"))))
```

### 2024-04-08 Sections/IGAS.qmd
render full in pdf and markdown
tweaks

The past century of research on human learning has produced ample evidence that although learners can improve at almost any task, such improvements are often specific to the trained task, with unreliable or even nonexistent transfer to novel tasks or conditions [@barnettWhenWhereWe2002; @dettermanCaseProsecutionTransfer1993]. Such transfer challenges are of noteworthy practical relevance, given that educators, trainers, and rehabilitators typically intend for their students to be able to apply what they have learned to new situations. It is therefore important to better understand the factors that influence transfer, and to develop cognitive models that can predict when transfer is likely to occur. The factor of interest to the present investigation is variation during training. Our experiments add to the longstanding empirical investigation of the controversial relationship between training variation, and subsequent transfer. We also offer a novel explanation for such results in the form of an instance-based model that accounts for the benefits of variation in simple terms of psychological similarity. We first review the relevant concepts and literature.

@@

While similarity-based models account for transfer by the degree of similarity between previous and new experiences, a largely separate body of research has focused on improving transfer by manipulating characteristics of the initial training stage. Such characteristics have included training difficulty, spacing, temporal order, feedback schedules, and the primary focus of the current work -- variability of training examples.

Research on the effects of varied training typically compares participants trained under constant, or minimal variability conditions to those trained from a variety of examples or conditions [@czyzVariabilityPracticeInformation2021; @soderstromLearningPerformanceIntegrative2015]. Varied training has been shown to influence learning in myriad domains including categorization of simple stimuli [@hahnEffectsCategoryDiversity2005; @maddoxStimulusRangeDiscontinuity2011; @posnerGenesisAbstractIdeas1968], complex categorization [@nosofskyModelguidedSearchOptimal2018], language learning [@jonesDensityDistinctivenessEarly2020; @perryLearnLocallyThink2010; @twomeyAllRightNoises2018; @wonnacottInputEffectsAcquisition2012], anagram completion [@goodeSuperiorityVariableRepeated2008], trajectory extrapolation [@fulvioTaskSpecificResponseStrategy2014], task switching [@sabahWhenLessMore2019], associative learning [@leeEvidentialDiversityIncreases2019], visual search [@georgeStimulusVariabilityTask2021; @gonzalezDiversityTrainingEnhances2011; @kelleyLearningAttendEffects2009], voice identity learning [@lavanEffectsHighVariability2019], simple motor learning [@braunMotorTaskVariation2009; @kerrSpecificVariedPractice1978; @rollerVariablePracticeLenses2001; @willeyLimitedGeneralizationVaried2018], sports training  [@greenPracticeVariabilityTransfer1995a; @northEffectConsistentVaried2019], and training on a complex video game [@seowTransferEffectsVaried2019].

@@

The experimental task was programmed in JavaScript, using packages from
the Phaser physics engine (https://phaser.io) and the jsPsych library
(de Leeuw, 2015). The stimuli, presented on a black background,
consisted of a circular blue ball -- controlled by the participant via
the mouse or trackpad cursor; a rectangular green target; a red
rectangular barrier located between the ball and the target; and an
orange square within which the participant could control the ball before
releasing it in a throw towards the target. Because the task was
administered online, the absolute distance between stimuli could vary
depending on the size of the computer monitor being used, but the
relative distance between the stimuli was held constant. Likewise, the
distance between the center of the target, and the training and testing
locations was scaled such that relative distances were preserved
regardless of screen size. For the sake of brevity, subsequent mentions
of this relative distance between stimuli, or the position where the
ball landed in relation to the center of the target, will be referred to
simply as distance. @fig-IGAS_Methods displays the layout of the task, as
it would appear to a participant at the start of a trial, with the ball
appearing in the center of the orange square. Using a mouse or trackpad,
participants click down on the ball to take control of the ball,
connecting the movement of the ball to the movement of the cursor.
Participants can then "wind up" the ball by dragging it (within the
confines of the orange square) and then launch the ball by releasing the
cursor. If the ball does not land on the target, participants are
presented with feedback in red text at the top right of the screen, on
how many units away they were from the center of the target. If the ball
was thrown outside of the boundary of the screen participants are given
feedback as to how far away from the target center the ball would have
been if it had continued its trajectory. If the ball strikes the barrier
(from the side or by landing on top), feedback is presented telling
participants to avoid hitting the barrier. If participants drag the ball
outside of the orange square before releasing it, the trial terminates,
and they are reminded to release the ball within the orange square. If
the ball lands on the target, feedback is presented in green text,
confirming that the target was hit, and presenting additional feedback
on how many units away the ball was from the exact center of the target.

[Link to abbrevaited example of
task](https://pcl.sitehost.iu.edu/tg/demos/igas_expt1_demo.html){target="_blank"}.

@@

Next, we compared the testing performance of constant and varied groups
from only positions that participants had not encountered during
training. Constant participants each had 5 novel positions, whereas
varied participants tested from 4 novel positions (400,625,675,900). We
first standardized performance within in each position, and then
averaged across positions. Here again, we found a significant effect of
condition (constant vs. varied): F(1,206)=4.30, p=.039, $\eta^{2}_G$ =
.02 .

```{r}
#| label: tbl-e2table2
#| tbl-cap: Testing performance from novel positions. Includes data only from positions that were not encountered during the training stage (e.g. excludes positions 500 and 800 for the varied group, and one of the six locations for each of the constant groups). Table presents Mean absolute deviations from the center of the target, and standard deviations in parenthesis.

@@

In the testing phase, our varied group significantly outperformed the
constant conditions in both a general comparison, and in an analysis
limited to novel throwing positions. The observed benefit of varied over
constant training echoes the findings of many previous visuomotor skill
learning studies that have continued to emerge since the introduction of
Schmidt's influential Schema Theory [@catalanoDistantTransferCoincident1984a; @chuaPracticeVariabilityPromotes2019; @goodwinEffectDifferentQuantities1998; @mccrackenTestSchemaTheory1977; @moxleySchemaVariabilityPractice1979; @newellVariabilityPracticeTransfer1976; @pigottMotorSchemaStructure1984; @rollerVariablePracticeLenses2001; @schmidtSchemaTheoryDiscrete1975; @willeyLongtermMotorLearning2018; @wrisbergVariabilityPracticeHypothesis1987; @wulfEffectTypePractice1991]. We also join a much smaller set of research to observe this
pattern in a computerized task [@seowTransferEffectsVaried2019]. One departure from
the experiment 1 findings concerns the pattern wherein the varied group
outperformed the constant group even from the training position of the
constant group, which was significant in experiment 1, but did not reach
significance in experiment 2. Although this pattern has been observed
elsewhere in the literature [@goodeSuperiorityVariableRepeated2008; @kerrSpecificVariedPractice1978], 
the overall evidence for this effect appears to be far weaker than for
the more general benefit of varied training in conditions novel to all
training groups.

# Computational Model

@@

Our experiments demonstrate a reliable benefit of varied training in a
simple projectile launching task. Such results were accounted for by an
instance-based model that assumes that varied training results in the
computation of a broader similarity-based generalization gradient.
Instance-based models augmented with this assumption may be a valuable
approach towards better understanding skill generalization and transfer.

>>>>

@@

@@

The experimental task was programmed in JavaScript, using packages from the Phaser physics engine (https://phaser.io) and the jsPsych library (de Leeuw, 2015). The stimuli, presented on a black background, consisted of a circular blue ball -- controlled by the participant via the mouse or trackpad cursor; a rectangular green target; a red rectangular barrier located between the ball and the target; and an orange square within which the participant could control the ball before releasing it in a throw towards the target. Because the task was administered online, the absolute distance between stimuli could vary depending on the size of the computer monitor being used, but the relative distance between the stimuli was held constant. Likewise, the distance between the center of the target, and the training and testing locations was scaled such that relative distances were preserved regardless of screen size. For the sake of brevity, subsequent mentions of this relative distance between stimuli, or the position where the ball landed in relation to the center of the target, will be referred to simply as distance. @fig-IGAS_Methods displays the layout of the task, as it would appear to a participant at the start of a trial, with the ball appearing in the center of the orange square. Using a mouse or trackpad, participants click down on the ball to take control of the ball, connecting the movement of the ball to the movement of the cursor. Participants can then "wind up" the ball by dragging it (within the confines of the orange square) and then launch the ball by releasing the cursor. If the ball does not land on the target, participants are presented with feedback in red text at the top right of the screen, on how many units away they were from the center of the target. If the ball was thrown outside of the boundary of the screen participants are given feedback as to how far away from the target center the ball would have been if it had continued its trajectory. If the ball strikes the barrier (from the side or by landing on top), feedback is presented telling participants to avoid hitting the barrier. If participants drag the ball outside of the orange square before releasing it, the trial terminates, and they are reminded to release the ball within the orange square. If the ball lands on the target, feedback is presented in green text, confirming that the target was hit, and presenting additional feedback on how many units away the ball was from the exact center of the target. [Link to abbrevaited example of task](https://pcl.sitehost.iu.edu/tg/demos/igas_expt1_demo.html){target="_blank"}.

@@

Next, we compared the testing performance of constant and varied groups from only positions that participants had not encountered during training. Constant participants each had 5 novel positions, whereas varied participants tested from 4 novel positions (400,625,675,900). We first standardized performance within in each position, and then averaged across positions. Here again, we found a significant effect of condition (constant vs. varied): F(1,206)=4.30, p=.039, $\eta^{2}_G$ = .02 .
```{r}
#| label: tbl-e2table2
#| tbl-cap: Testing performance from novel positions. Includes data only from positions that were not encountered during the training stage (e.g. excludes positions 500 and 800 for the varied group, and one of the six locations for each of the constant groups). Table presents Mean absolute deviations from the center of the target, and standard deviations in parenthesis.

@@

In the testing phase, our varied group significantly outperformed the constant conditions in both a general comparison, and in an analysis limited to novel throwing positions. The observed benefit of varied over constant training echoes the findings of many previous visuomotor skill learning studies that have continued to emerge since the introduction of Schmidt's influential Schema Theory [@catalanoDistantTransferCoincident1984a; @chuaPracticeVariabilityPromotes2019; @goodwinEffectDifferentQuantities1998; @mccrackenTestSchemaTheory1977; @moxleySchemaVariabilityPractice1979; @newellVariabilityPracticeTransfer1976; @pigottMotorSchemaStructure1984; @rollerVariablePracticeLenses2001; @schmidtSchemaTheoryDiscrete1975; @willeyLongtermMotorLearning2018; @wrisbergVariabilityPracticeHypothesis1987; @wulfEffectTypePractice1991]. We also join a much smaller set of research to observe this pattern in a computerized task [@seowTransferEffectsVaried2019]. One departure from the experiment 1 findings concerns the pattern wherein the varied group outperformed the constant group even from the training position of the constant group, which was significant in experiment 1, but did not reach significance in experiment 2. Although this pattern has been observed elsewhere in the literature [@goodeSuperiorityVariableRepeated2008; @kerrSpecificVariedPractice1978], the overall evidence for this effect appears to be far weaker than for the more general benefit of varied training in conditions novel to all training groups.
# Computational Model

@@

Our experiments add to the longstanding body of research investigating  the effect of training variability on learning and generalization. Both experiments demonstrate a reliable benefit of varied training over constant training in a projectile launching task. We also introduce the Instance-based Generalization with Adaptive Similarity (IGAS) as a novel explanation for the effect of variability on generalization. The key assumption made by IGAS is that the amount of variation encountered during training influences the steepness of the generalization gradient. Instance-based models augmented with this assumption may be a valuable approach towards better understanding skill generalization and transfer.

### 2024-04-08 Sections/Introduction.qmd
tweaks

# Variability and Generalization

The factors that influence the generalization of learning are of considerable interest to both researchers exploring the human learning system and practitioners aiming to enhance the effectiveness of educational and training interventions. The present effort will focus specifically on the role of variability in the learning input. Variability manipulations typically regulate either the number of distinct instances presented to learners during training, or the dispersion of these instances. Such manipulations have been empirically demonstrated to affect subsequent generalization performance. This essay will offer an in-depth review of the extant literature on the influence of variability, spanning multiple relevant domains.

@@

Studies investigating the "benefits of variability" hypothesis usually assign participants to either a constant or varied group for the training stage of the experiment. Then, subjects in both groups complete an identical testing stage which often consists items/conditions seen during training, and novel items/conditions. If the varied group performs better in the testing stage, this is taken for evidence of the benefits of variability hypothesis. Even within this relatively straightforward between-groups design, researchers must navigate several crucial methodological choices, highlighted below:

1)  Variables Subject to Variation. In multidimensional tasks, researchers have the option to vary numerous variables. The experimenters must decide the specific dimension(s) across which variation will occur. For instance, in a projectile throwing accuracy task -- researchers might vary the distance from the target, the size of the target, the weight of the projectile. They might also vary a contextual variable not directly relevant to the task, but which will still be encoded by the subject on a trial by trial basis, e.g. the background color.

2)  Magnitude of Variation, relative to the control condition. The simplest comparison would be to compare a constant group who trains with 1 example/condition, against a varied group that trains from 2 examples/conditions. However, it is not uncommon in the literature for the varied condition to train from 3 or 4 conditions. For example, @catalanoDistantTransferCoincident1984a train varied subjects from 4 different velocities in their coincident timing task, and [@goodeSuperiorityVariableRepeated2008] have varied subjects' practice with 3 different variants (i.e. different letter scrambles of the same word) of an anagram for a given word, while their constant participants view the same variant 3 different times. Alternatively, rather than a constant vs. varied comparison, subjects in all conditions might experience a variety of training items, but with one group experiencing a greater number of unique items [@nosofskyModelguidedSearchOptimal2018].

3)  Locations within Task-Space. For tasks in which the stimuli or conditions fall within a continuous metric space, the experimenter must decide whether the varied instances are relatively close together (e.g. throwing a ball from a distance of 4 feet and 5 feet), or far apart (throwing from 4 feet and 20 feet). Spreading the varied training items further apart may be beneficial in terms of providing a more representative sample of the task space to the learner, however large distances may also result in significant differences in difficulty between the training examples, which can be a common confound in variability studies.

4)  Proximity of Testing to Training Conditions. Intuitively, the fairest form of comparison is to include testing conditions that are of an equivalent distance from both the varied and constant groups. However researchers might also attempt to demonstrate the benefits of variation as being sufficiently powerful to outperform constant training, even in cases where the constant group trained from a closer proximity to the testing conditions, or whose training conditions are identical to the testing conditions [@goodeSuperiorityVariableRepeated2008; @kerrSpecificVariedPractice1978].

@@

In the category learning literature, the constant vs. varied comparison is much less suitable. Instead, researchers tend to compare a condition with many repetitions of a few items against condition with fewer repetitions of a wider array of exemplars. Much of the earlier work in this sub-area trained subjects on artificial categories, such as dot patterns [@homaCategoryBreadthAbstraction1976; @posnerGenesisAbstractIdeas1968], where more varied or distorted training examples were often shown to produce superior generalization when categorizing novel exemplars. More recently, researchers have also begun to utilize more realistic stimuli in their experiments. @wahlheimMetacognitiveJudgmentsRepetition2012 conducted one such study. In a within-participants design, participants were trained on bird categories with either high repetitions of a few exemplars, or few repetitions of many exemplars. Across four different experiments, which were conducted to address an unrelated question on metacognitive judgements, the researchers consistently found that participants generalized better to novel species following training with more unique exemplars (i.e. higher variability), while high repetition training produced significantly better performance categorizing the specific species they had trained on. A variability advantage was also found in the relatively complex domain of rock categorization [@nosofskyTestsExemplarmemoryModel2018]. For 10 different rock categories, participants were trained with either many repetitions of 3 unique examples of each category, or few repetitions of 9 unique examples, with an equal number of total training trials in each group (the design also included 2 other conditions less amenable to considering the impact of variation). The high-variability group, trained with 9 unique examples, showed significantly better generalization performance than the other conditions. Moreover, the pattern of results in this study could be nicely accounted for by an extended version of the Generalized Context Model.

@@

Of course, the effects of varied training, and different training sequences, are likely to be far more complex than simply more varied training being better than less, or random training being better than blocked. Null effects of both manipulations have been reported [see @magillReviewContextualInterference1990; @vanrossumSchmidtSchemaTheory1990 for reviews], and a variety of moderators have emerged. In one of the earlier examples of the complex relationship between study sequence and learning [@delreyEffectsContextualInterference1982], experimenters recruited participants who self-reported either large amounts, or very little experience with athletic actives, and then trained participants on a coincident timing task under with either a single constant training velocity, or with four training velocities under either blocked, or random training sequence conditions - resulting in six experimental conditions: (athlete vs. non-athlete) x (constant vs. blocked vs. random training). Athlete participants had superior performance during training, regardless of sequence condition, and training performance was superior for all subjects in the constant group, followed by blocked training in the middle, and then random training resulting in the worst training performance. Of greater interest is the pattern of testing results for novel transfer conditions. Among the athlete-participants, transfer performance was best for those who received random training, followed by blocked, and then constant training. Non-athletes showed the opposite pattern, with superior performance for those who had constant training. A similar pattern was later observed in a golf-putting training study, wherein participants who had some prior golf experience benefited most from random-sequenced training, and participants with no golf experience benefited most from blocked training  [@guadagnoliRelationshipContextualInterference1999]. More recently, the same pattern was observed in the concept learning literature [@braithwaiteEffectsVariationPrior2015 expt 1.]. This study trained participants on a mathematical concept and found that participants who self-reported some prior experience with the concept improved more from pre-test to post-test after training with varied examples, while participants who reported no prior experience showed greater gains following training with highly similar examples.

>>>>

# Introduction

The past century of research on human learning has produced ample evidence that although learners can improve at almost any task, such improvements are often specific to the trained task, with unreliable or even nonexistent transfer or generalization to novel tasks or conditions [@barnettWhenWhereWe2002; @dettermanCaseProsecutionTransfer1993]. Such generalization challenges are of noteworthy practical relevance, given that educators, trainers, and rehabilitators typically intend for their students to be able to apply what they have learned to new situations. It is therefore important to better understand the factors that influence generalization, and to develop cognitive models that can predict when generalization is likely to occur. Such characteristics have included training difficulty, spacing, temporal order, feedback schedules, and the primary focus of the current work - the variability of training examples.

@@

Varied training has been shown to influence learning in wide array of different tasks and domains, including categorization [@hahnEffectsCategoryDiversity2005; @maddoxStimulusRangeDiscontinuity2011; @posnerGenesisAbstractIdeas1968; @nosofskyModelguidedSearchOptimal2019], language learning [@jonesDensityDistinctivenessEarly2020; @perryLearnLocallyThink2010; @twomeyAllRightNoises2018; @wonnacottInputEffectsAcquisition2012], anagram completion [@goodeSuperiorityVariableRepeated2008], trajectory extrapolation [@fulvioTaskSpecificResponseStrategy2014], task switching [@sabahWhenLessMore2019], associative learning [@leeEvidentialDiversityIncreases2019], visual search [@georgeStimulusVariabilityTask2021; @gonzalezDiversityTrainingEnhances2011; @kelleyLearningAttendEffects2009], voice identity learning [@lavanEffectsHighVariability2019], simple motor learning [@braunMotorTaskVariation2009; @kerrSpecificVariedPractice1978; @rollerVariablePracticeLenses2001; @willeyLimitedGeneralizationVaried2018], sports training  [@greenPracticeVariabilityTransfer1995a; @northEffectConsistentVaried2019], and training on a complex video game [@seowTransferEffectsVaried2019].

Research on the effects of varied training typically compares a group of participants trained with minimal variability of examples against a group with greater variability [@czyzVariabilityPracticeInformation2021; @ravivHowVariabilityShapes2022].

@@

In the category learning literature, the constant vs. varied comparison is much less suitable. Instead, researchers tend to compare a condition with many repetitions of a few items against condition with fewer repetitions of a wider array of exemplars. Much of the earlier work in this sub-area trained subjects on artificial categories, such as dot patterns [@homaCategoryBreadthAbstraction1976; @posnerGenesisAbstractIdeas1968], where more varied or distorted training examples were often shown to produce superior generalization when categorizing novel exemplars. More recently, researchers have also begun to utilize more realistic stimuli in their experiments. @wahlheimMetacognitiveJudgmentsRepetition2012 conducted one such study. In a within-participants design, participants were trained on bird categories with either high repetitions of a few exemplars, or few repetitions of many exemplars. Across four different experiments, which were conducted to address an unrelated question on metacognitive judgements, the researchers consistently found that participants generalized better to novel species following training with more unique exemplars (i.e. higher variability), while high repetition training produced significantly better performance categorizing the specific species they had trained on. A variability advantage was also found in the relatively complex domain of rock categorization [@nosofskyModelguidedSearchOptimal2019]. For 10 different rock categories, participants were trained with either many repetitions of 3 unique examples of each category, or few repetitions of 9 unique examples, with an equal number of total training trials in each group (the design also included 2 other conditions less amenable to considering the impact of variation). The high-variability group, trained with 9 unique examples, showed significantly better generalization performance than the other conditions. Moreover, the pattern of results in this study could be nicely accounted for by an extended version of the Generalized Context Model.

@@

Of course, the effects of varied training, and different training sequences, are likely to be far more complex than simply more varied training being better than less, or random training being better than blocked. Null effects of both manipulations have been reported [see @magillReviewContextualInterference1990; @vanrossumSchmidtSchemaTheory1990 for reviews], and a variety of moderators have emerged. In one of the earlier examples of the complex relationship between study sequence and learning [@delreyEffectsContextualInterference1982], experimenters recruited participants who self-reported either large amounts, or very little experience with athletic activities, and then trained participants on a coincident timing task under with either a single constant training velocity, or with four training velocities under either blocked, or random training sequence conditions - resulting in six experimental conditions: (athlete vs. non-athlete) x (constant vs. blocked vs. random training). Athlete participants had superior performance during training, regardless of sequence condition, and training performance was superior for all subjects in the constant group, followed by blocked training in the middle, and then random training resulting in the worst training performance. Of greater interest is the pattern of testing results for novel transfer conditions. Among the athlete-participants, transfer performance was best for those who received random training, followed by blocked, and then constant training. Non-athletes showed the opposite pattern, with superior performance for those who had constant training. A similar pattern was later observed in a golf-putting training study, wherein participants who had some prior golf experience benefited most from random-sequenced training, and participants with no golf experience benefited most from blocked training  [@guadagnoliRelationshipContextualInterference1999]. More recently, the same pattern was observed in the concept learning literature [@braithwaiteEffectsVariationPrior2015 expt 1.]. This study trained participants on a mathematical concept and found that participants who self-reported some prior experience with the concept improved more from pre-test to post-test after training with varied examples, while participants who reported no prior experience showed greater gains following training with highly similar examples.
